---
title: "The most severe weather events in the United States in terms of the population health and the economic consequences"
author: "Marcin Zakidalski"
date: "23 05 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The synopsis

This analysis is going to answer the following two questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

The data used in this analysis comes from the following source: <https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2>.

## Data Processing

R packages used in the analysis:

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(readr)
library(ggplot2)
library(dplyr)
library(stringr)
```

From the data set we are interested only in the following columns:

* STATE - code abbreviating the state
* BGN_DATE, END_DATE - event start/end dates
* PROPDMG / CROPDMG - describes the damage to the property/crop
* PROPDMGEXP / CROPDMGEXP - multipliers to the previous values (k/K for 1000, m/M for 1 000 000, b/B for 1 000 000 000).                                

Data loading & column selection (we assume that the download dataset is present in the working folder):

```{r, echo=TRUE, warning=FALSE, message=FALSE}

all_weather_data <- read_csv("repdata_data_StormData.csv.bz2")
filtered_cols_data <- all_weather_data %>% select(STATE, BGN_DATE, END_DATE, 
                                                  EVTYPE,
                                                  PROPDMG, PROPDMGEXP, CROPDMG, 
                                                  CROPDMGEXP, INJURIES, FATALITIES)
```

The filtered_cols_data dataframe contains the rows "Summary..." and "?" ENVTYPEs  which should be deleted. Additionally event names are transformed to the uppercase:

```{r, echo=TRUE, warning=FALSE, message=FALSE}
head(filtered_cols_data %>% filter (str_detect(EVTYPE, "[Ss][Uu][Mm][Mm][Aa][Rr][Yy]")) %>% select(EVTYPE))
filtered_cols_data <- filtered_cols_data %>% filter (!str_detect(EVTYPE, "[Ss][Uu][Mm][Mm][Aa][Rr][Yy]")) %>% 
                      filter (!str_detect(EVTYPE, "\\?")) %>% 
                      mutate(EVTYPE = toupper(EVTYPE))
```

According to the documentation of the dataset the COLNAME*EXP columns contain the multiplier
for the COLNAME values. This data contains not only tidy values:

```{r, echo=TRUE, warning=FALSE, message=FALSE}

print( unique(filtered_cols_data$PROPDMGEXP))
print( unique(filtered_cols_data$CROPDMGEXP))
```

The analysis assumption is that:

* allowed multipliers are B, M (which is equivalent to m), K
* all other values (including NAs) are equivalent to 1.0 multiplier

The ALL_* columns will contain the real value (in $) of damages:

```{r, echo=TRUE, warning=FALSE, message=FALSE}
filtered_cols_data <- filtered_cols_data %>%
                          mutate(ALL_PROPDMG = case_when( endsWith(PROPDMGEXP,"B")  ~ 1000000000.0*PROPDMG,
                                                          endsWith(PROPDMGEXP, "M") ~ 1000000.0*PROPDMG,
                                                          endsWith(PROPDMGEXP, "m") ~ 1000000.0*PROPDMG,
                                                          endsWith(PROPDMGEXP, "K") ~ 1000.0*PROPDMG,
                                                          TRUE ~ 1.0*PROPDMG))

# there are no multipliers for the crop-related damage
filtered_cols_data <- filtered_cols_data %>% mutate(ALL_CROPDMG = CROPDMG)
```

We can briefly verify our transformations are correct:
```{r, echo=TRUE, warning=FALSE, message=FALSE}
head(filtered_cols_data %>% filter(PROPDMG > 0) %>% select(-STATE, -BGN_DATE, -END_DATE))
head(filtered_cols_data %>% filter(CROPDMG > 0) %>% select(-STATE, -BGN_DATE, -END_DATE))
```

Finally, we prepare the summary dataframe to work further on:

```{r, echo=TRUE, warning=FALSE, message=FALSE}
events_summary <- filtered_cols_data %>% select(EVTYPE, INJURIES, FATALITIES, ALL_PROPDMG, ALL_CROPDMG) %>%
                                         mutate(ALL_DMG = ALL_PROPDMG + ALL_CROPDMG) %>%
                                         select(-ALL_PROPDMG, -ALL_CROPDMG) %>%
                                         group_by(EVTYPE) %>%
                                         summarise(across(where(is.numeric), sum))
events_summary$EVTYPE <- as.factor(events_summary$EVTYPE)
head(events_summary)
```


## Results

Here are the answers to the questions stated in the synopsis:
1. We define the events which are most harmful with respect to population health as fulfilling any of the following conditions:
   
   
* top five events according to the cumulative fatalities across all data from the dataset

```{r, echo=TRUE, warning=FALSE, message=FALSE}
answer_1a <- events_summary %>% arrange(desc(FATALITIES)) %>% select (EVTYPE, FATALITIES)
head(answer_1a, n=5)
```   
   
   
* top five events according to the cumulative injuries across all data from the dataset
```{r, echo=TRUE, warning=FALSE, message=FALSE}
answer_1b <- events_summary %>% arrange(desc(INJURIES)) %>% select (EVTYPE, INJURIES)
head(answer_1b, n=5)
```   
   
2. We define the events with the top economic influence as those causing top 5 cumulative damage costs across all the dataset
```{r, echo=TRUE, warning=FALSE, message=FALSE}
answer_2 <- events_summary %>% arrange(desc(ALL_DMG)) %>% select (EVTYPE, ALL_DMG)
head(answer_2, n=5)
```   

We can also see a histogram of those costs:
```{r, echo=TRUE, warning=FALSE, message=FALSE}
answer_2 <- answer_2 %>% top_n(ALL_DMG, n=5)
ggplot(answer_2, aes(x=EVTYPE, y=ALL_DMG)) +geom_bar(stat="identity")+
      xlab("Event type") + ylab("Cumulative costs of events")
```